---
title: "Target Vertices Processing"
author: "Guillaume Guénard"
date: "2025-01-28"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(MPSEM)
```

## Examplary graph

```{r}
gr_ex2 <- readRDS("gr_ex2.rds")
plot(gr_ex2, cex.min=3.5, cex.lab=0.7)
```
A `r nrow(gr_ex2)`-vertex `r nedge(gr_ex2)`-edge exemplary graph used for this
discussion.
---

## Target calculation

Cases outlined in this document are for a vertex $V_T$ (target) preceded by
$n_p$ vertices $V^{(P)}_1$, $V^{(P)}_2$, ..., $V^{(P)}_{n_f}$, and followed by
$n_f$ vertices $V^{(F)}_1$, $V^{(F)}_2$, ..., $V^{(F)}_{n_f}$. The edges
linking $V_{P_i}$ and $V_T$ are referred to as $E^{(P)}_1$, $E^{(P)}_2$, ...,
$E^{(P)}_{n_p}$ and their associated distance are $D^{(P)}_1$, $D^{(P)}_2$,
..., $D^{(P)}_{n_p}$, whereas that linking $V_T$ to $V_{F_i}$ are referred to
as $E^{(F)}_1$, $E^{(F)}_2$, ..., $E^{(F)}_{n_f}$, and their associated
distances are $D^{(F)}_1$, $D^{(F)}_2$, ..., $D^{(F)}_{n_f}$. From these
definitions it follows that a vertex is:

* convergent when $n_p > 1$ or non-convergent when $n_p = 1$,
* divergent when $n_f > 1$ or non-divergent when $n_f = 1$, and
* terminal when $n_f = 0$ or non-terminal when $n_f > 0$.

When evaluating the loading of a target vertex on the phylogenetic
eigenfunctions, the later may remain or be discarded, depending on the
dependence of other vertices upon it. In PEM, not all vertices have values
associated with them. Vertices with associated values are indicated using a
logical variable called `species`, which is `TRUE` for vertices with values and
`FALSE` otherwise. Therefore, only those vertices with no upward dependent
vertices that are not marked as "species" can be readily discarded unless they
are identified as target species.






Vertices whose removal would affect other vertices



that have other vertices above them cannot
be removed, whereas vertices whose removal affects no 


Non-convergent and non-divergent (**NCND**) vertices are
always removable (both terminal and non-terminal), whereas convergent or
divergent vertices are not. Non-convergent, but divergent vertices may become
non-divergent upon removal of other vertices located after them, and thus some
non-convergent vertices may eventually become removable. There is no way for
convergent vertices to become removable.

When a target vertex is not removable, its `$species` nodal value is set to
`FALSE` and its influence coordinates are used to calculate the loadings. A
removed vertex does no longer belongs to the graph, but maintains an
ancestral reference to one of the graph's vertex, or to an intersection point
within one the the graph's edge. Two possibilities exist. A terminal $V^{(T)}$
can be removed along with $E^{(P)}_1$ and its influence coordinates are that of
$V^{(P)}_1$ with a distance from the latest ancestor of $D_{la} = D^{(P)}_1$.
For a non-terminal $V^{(T)}$, $E^{(F)}_1$ is deleted and $E^{(P)}_1$ is extended
to reach $V^{(F)}_1$ (denoted $E^{(P-F)}_1$, its new distance becomes
$D^{(P-F)}_1 = D^{(P)}_1 + D^{(F)}_1$) and its coordinates are located on
$E^{(P)}_1$ at distance $D^{(P)}_1$ with a distance from the latest ancestor of
$D_{la} = 0$. 




An ancestral reference
to an edge thus also involves a value of distance between the origin of that
edge and the intersection point.

A vertex can be removed if and only if no other vertices is linked to it. Since
convergent and divergent vertices have other vertices before and after them,
they cannot be removed. Non-convergent vertices that with $n_f \leq 1$ such as
terminal and intermediary vertices can be removed. Also, an erstwhile divergent
vertex that have become a terminal or intermediary vertex following the removal
of one or more of its tributaries can be removed.

Algorithm:

1. Check 1 -- verify that all the targets are marked as species, stop with an
   error message if any is not thus marked.

2. Simplification 1 -- going through all the vertices; verify if any vertex that
   is marked as no being a species can be removed readily. When one is found, it
   is removed and the procedure is restarted from the beginning. This step ends
   when going through all the vertices that are not marked as a species and
   finding none that can be removed. This step is optional, yet recommended, and
   is accomplished using an external function.
   
3. From the set of all target vertices, verify if there is a terminal vertex
   among them if there is one, go to step 5, otherwise go to 4.
   
4. From the set of all target vertices, verify if there is an intermediary
   vertex among them, if there is one, go to step 5, otherwise go to 6.
   
5. Remove the vertex earmarked for removal in step 3 (terminal vertex) or 4
   (intermediary vertex), then go to step 3.
   
   **Simplification 2**
   
6. From the set of all non-target vertices, verify if there is a terminal
   vertex among them, if there is one, go to step 8, otherwise, go to 7.
   
7. From the set of all non-target vertices, verify if there is an intermediary
   vertex among them, if there is one, go to step 8, otherwise, to to 9.

8. Remove the vertex earmarked for removal in step 6 (terminal vertex) or 7
   (intermediary vertex), then go to step 6.
   
   **Non removables**
   
9. If there are any target vertex remaining, they cannot be removed.






  
  Pass 2: Identify the terminal vertices 


* Remove the terminal targets
* When there is no more 






C, whereas . On the other hand, strictly divergent
(ie. a non-convergent) vertex may 

divergent vertices may become an
intermediary of a terminal vertex upon removal one or more of its tributaries.
This situation calls for an iterative approach for target loading calculation.



The root and convergent vertices cannot be removed 


The calculation of the loadings needs to be organized in such a way that
terminal and intermediate vertices are treated first 



either a surrogate of an other vertex under which it
used to located or occurring at some location within an edge.  (with some
ancestral distance)



When a vertex is collapsed, it either becomes a surrogate of an other vertex

or
it is 






### Case : Intermediary vertex

The latter is arguably simplest case; if one wants to estimate the loading for
an intermediary vertex, vertices $V^{(P)}_1$ and $V^{(F)}_1$ can simply be
connected directly using a new edge with length $D^{(P)}_1 + D^{(F)}_1$. The
coordinates are that of $V^{(P)}_1$ at a distance of $D^{(P)}_1$ along the new
edge and a latest ancestor distance $D_{anc}=0$ is applied because the vertex is
indeed exactly within the edge. This is the case of strings of any number of
intermediary vertices.
