---
title: "Simulating phylogenetic networks using MPSEM"
author: "Guillaume Guénard"
date: "`r format(Sys.time(), '%Y-%m-%d')`"
output:
  rmarkdown::html_vignette:
    toc: true
    toc_depth: 3
bibliography: ../inst/bib/Simulating_phylogenetic_network.bib
vignette: >
  %\VignetteIndexEntry{Simulating phylogenetic networks using MPSEM}
  %\VignetteEngine{knitr::rmarkdown}
  \usepackage[utf8]{inputenc}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
hook_output <- knitr::knit_hooks$get("output")
knitr::knit_hooks$set(output = function(x, options) {
  if (!is.null(n <- options$out.lines)) {
    x <- xfun::split_lines(x)
    if (length(x) > n) {
      # truncate the output
      x <- c(head(x, n), "....\n")
    }
    x <- paste(x, collapse = "\n")
  }
  hook_output(x, options)
})
##
### Load packages here:
##
### Figure counter:
(
  function() {
    log <- list(
      labels = character(),
      captions = character()
    )
    list(
      register = function(label, caption) {
        log$labels <<- c(log$labels, label)
        log$captions <<- c(log$captions, caption)
        invisible(NULL)
      },
      getNumber = function(label) {
        which(log$labels == label)
      },
      getCaption = function(label) {
        a <- which(log$labels == label)
        cap <- log$captions[a]
        cat(sprintf("Fig. %d. %s\n\n---\n",a,cap))
        invisible(NULL)
      }
    )
  }
)() -> figCounter
```

Package version: `r packageVersion("MPSEM")`


# Introduction

Package **MPSEM** has functions to simulate evolutionary graphs, instances of
which can be simple a linear evolutionary sequence, a tree (i.e., phylogeny), or
reticulated evolutionary graphs (i.e., networks) involving hybridization between
branches (i.e., lateral gene transfer events). Species evolution has
historically been modeled as a tree. In a tree, any given species has a single
ancestor. This model only allows for longitudinal gene transfer, disregarding
lateral gene transfer processes. The simplicity of phylogeny (i.e., the
representation of evolution as a tree) overlooks any lateral evolutionary
processes such as recombination (new genomes are reconstituted through the
sorting and crossing-over of existing genes), hybridization (reproductive events
involving relatively distant parents), the exchange of independent transferable
elements (e.g., plasmides), and so on. Representing evolution as a reticulated
graph (as opposed to a tree), allows for the representation of lateral gene
transfer, at the price of a greater complexity.

Here, we will explain how to use MPSEM to generate random graphs and simulate
the evolution of DNA sequences and traits along them. The data thus generated,
for known graphs, and under controlled processes, are then used for various
purposes. For instance, they can be used in testing the performance of
quantitative method, for developing supervised and non-supervised machine
learning methods for the estimation of evolutionary distances, graph structure,
and so on.


# Usage

Before anything else, package **MPSEM** has to be loaded as follows:

```{r}
library(MPSEM)
```

## Molecular (DNA) evolution simulator

**MPSEM** simulates DNA sequence evolution as a Markov process whereby the
four nucleotides (**A**, **G**, **C**, or **T**) shift from one value to another
with probabilities defined by a shift rate matrix $\mathbf{Q}$, which is defined
as follows:

$$
\mathbf{Q} = \begin{bmatrix}
-\Sigma_\mathrm{A} & r_\mathrm{AG} \pi_\mathrm{G} & r_\mathrm{AC} \pi_\mathrm{C} & r_\mathrm{AT} \pi_\mathrm{T} \\
r_\mathrm{GA} \pi_\mathrm{A} & -\Sigma_\mathrm{G} & r_\mathrm{GC} \pi_\mathrm{C} & r_\mathrm{GT} \pi_\mathrm{T} \\
r_\mathrm{CA} \pi_\mathrm{A} & r_\mathrm{CG} \pi_\mathrm{G} & -\Sigma_\mathrm{C} & r_\mathrm{CT} \pi_\mathrm{T} \\
r_\mathrm{TA} \pi_\mathrm{A} & r_\mathrm{TG} \pi_\mathrm{G} & r_\mathrm{TC} \pi_\mathrm{C} & -\Sigma_\mathrm{T}
\end{bmatrix},
$$

where $r_{x,y}$ is the raw shift rate from state $x$ to state $y$,
$\pi_\mathrm{x}$ is the equilibrium relative frequency of DNA base $x$, and
$\Sigma_x$ is the sum of the off-diagonal elements on row $x$ (i.e.,
$\Sigma_\mathrm{A} = r_\mathrm{AG} \pi_\mathrm{G} + r_\mathrm{AC} \pi_\mathrm{C} + r_\mathrm{AT} \pi_\mathrm{T}$,
$\Sigma_\mathrm{G} = r_\mathrm{GA} \pi_\mathrm{G} + r_\mathrm{GC} \pi_\mathrm{C} + r_\mathrm{GT} \pi_\mathrm{T}$
and so on). The sum of nucleotide equilibrium (relative) frequencies is $1$ and
the values of the $12$ $r_\mathrm{xy}$ coefficients involved in shifts among the
nucleotides are standardized to an average of $1$. These two properties ensure 
the standardization of $\mathbf{Q}$. **MPSEM** also simulates the insertion and
deletion of nucleotides. That simulation is carried out by adding gaps
(${\text -}$) to the sequence and allowing each gap to shift into a nucleotide,
thereby simulating an insertion, or each nucleotide to turn into a gap, thereby
simulating a deletion. The shift rate matrix used to simulate that process is
defined as follows:

$$
\mathbf{Q^*} = \begin{bmatrix}
-\Sigma_\mathrm{\text -} & r_\mathrm{{\text -} N} (1 - \pi_{\text -}) \\
r_\mathrm{N {\text -}} \pi_{\text -} & \mathbf{Q} (1 - \pi_{\text -})
\end{bmatrix},
$$

where $r_\mathrm{{\text -} N}$ and $r_\mathrm{N {\text -}}$ are the raw
nucleotide insertion and deletion rates, respectively, and $\pi_{\text -}$ is
the equilibrium relative gap frequency in the sequence. The resulting
$5 \times 5$ shift matrix is defined as follows:

$$
\mathbf{Q^*} = \begin{bmatrix}
-\Sigma_\mathrm{{\text -}} & r_\mathrm{{\text -} A} \pi_\mathrm{A} (1 - \pi_{\text -}) & r_\mathrm{{\text -} G} \pi_\mathrm{G} (1 - \pi_{\text -}) & r_\mathrm{{\text -} C} \pi_\mathrm{C} (1 - \pi_{\text -}) & r_\mathrm{{\text -} T} \pi_\mathrm{T} (1 - \pi_{\text -}) \\
r_\mathrm{A {\text -}} \pi_{\text -} & -\Sigma_\mathrm{A} & r_\mathrm{AG} \pi_\mathrm{G} (1 - \pi_{\text -}) & r_\mathrm{AC} \pi_\mathrm{C} (1 - \pi_{\text -}) & r_\mathrm{AT} \pi_\mathrm{T} (1 - \pi_{\text -}) \\
r_\mathrm{G {\text -}} \pi_{\text -} & r_\mathrm{GA} \pi_\mathrm{A} (1 - \pi_{\text -}) & -\Sigma_\mathrm{G} & r_\mathrm{GC} \pi_\mathrm{C} (1 - \pi_{\text -}) & r_\mathrm{GT} \pi_\mathrm{T} (1 - \pi_{\text -}) \\
r_\mathrm{C {\text -}} \pi_{\text -} & r_\mathrm{CA} \pi_\mathrm{A} (1 - \pi_{\text -}) & r_\mathrm{CG} \pi_\mathrm{G} (1 - \pi_{\text -}) & -\Sigma_\mathrm{C} & r_\mathrm{CT} \pi_\mathrm{T} (1 - \pi_{\text -}) \\
r_\mathrm{T {\text -}} \pi_{\text -} & r_\mathrm{TA} \pi_\mathrm{A} (1 - \pi_{\text -}) & r_\mathrm{TG} \pi_\mathrm{G} (1 - \pi_{\text -}) & r_\mathrm{TC} \pi_\mathrm{C} (1 - \pi_{\text -}) & -\Sigma_\mathrm{T}
\end{bmatrix},
$$

where $r_\mathrm{{\text -} x}$ and $r_\mathrm{x {\text -}}$ are the (raw) rates
of nucleotide $\mathrm{x}$ insertion and deletion, respectively. The sum of the
rows of $\mathbf{Q^*}$ is $0$ by construct. The probability density matrix
$\mathbf{P}$ is obtained from the shift rate matrix $\mathbf{Q^*}$ as follows:

$$
\mathbf{P} = e^{\rho t \mathbf{Q^*}}
$$

where $e$ is Euler's number, $t$ is an amount of time, and $\rho$ is the mean
shift rate over all possible shifts, with the rows of $\mathbf{P}$ indeed
summing to $1$. **MPSEM** implements eight common DNA evolution models, all
of which being time-reversible ($r_\mathrm{xy} = r_\mathrm{yx}$ for all
nucleotides $\mathrm{x}$ and $\mathrm{y}$). This task is accomplished by
function `DNArate()`, which is called as follows:

```{r}
Q <- DNArate(model="JC69", piGap=0.3, insertionRate=0.1, deletionRate=0.1)
Q
```

This shift rate matrix corresponds to the @JC69model DNA evolution model, with
and equilibrium gap frequency of $30\%$ and insertion and deletion rates of
$0.1$. **MPSEM** implements molecular evolution models using function
`molEvolSim()`, which is given the shift rate matrix as its first argument, in
addition to values of time step ($t$, argument `step`) and mean shift rate
($\rho$, argument `rho`). A DNA molecular simulator is implemented as follows:

```{r}
em1 <- molEvolSim(Q, step = 1, rho = 5)
```

The function returns a list with three functions: `$recalculate(t, rho)`
recalculates the simulator for new values of $t$ and $\rho$, `$evolve(N)` take a
nucleotide value (`raw`) and returns a nucleotide value drawn from the posterior
probability distribution of the molecular evolution model, whereas `$getMt()`
returns the shift probability matrix of the simulator. For instance:

```{r}
em1$getMt()
```

is the shift probability matrix for a time step of $1$ and an mean shift rate
of $5$, if we recalculate the simulator for a time step of $0.1$:

```{r}
em1$recalculate(0.1, 5)
em1$getMt()
```

the shift probabilities decrease as less time is allowed for a change to take
place, whereas if we increase the time step to $5$:

```{r}
em1$recalculate(5, 5)
em1$getMt()
```

the shift probabilities increase as more time is allowed for a change to take
place.

> In `R`, it is more efficient when playing with single characters of a
> character string to work with a `raw` vector. A raw vector enables one to
> access single character with regular subset operators (`[]`) rather than using
> string operators. Function `rawToChar()` is used to turn a raw vector into a
> string.

In the following examples, let us create start with an `"A"` at time $0$ and
evolve it for $100$ time steps of $1$:

```{r}
set.seed(1324987)                 ## To obtain a consistent example
em1$recalculate(1, 5)             ## Recalculate for a time step of 1
s <- raw(101)                     ## For storage
s[1] <- charToRaw("A")            ## Initialize with and "A"
for(i in 2:101)                   ## Simulation loop:
  s[i] <- em1$evolve(s[i - 1]) 
rawToChar(s)                      ## Show the sequence
```

This example show multiple cases of nucleotide conservation (two or more
repetitions of the same value), shifts, insertion deletion and insertions.

The assumption that all sites of a DNA sequence evolve at the same rate
not supported in practice. To obtain a credible simulation, it is more suitable
to generate multiple values and assign them to individual sites. **MPSEM** has a
function called `drawEvolRate()` which draws random values from a gamma
distributions as follows:

```{r}
er <- drawEvolRate(NN=70, gamma.shape=3, gamma.scale=2e-2)
er
```

where `NN` is the number of sites of the sequence, `gamma.shape` is the shape
parameter of the gamma distribution from which the values are drawn, and
`gamma.scale` is the latter's scale parameter. **MPSEM** also has a function
called `drawDNASequence()` to help in drawing initial random sequences from
which to initiate simulation process. It is called as follows:

```{r}
set.seed(13987463)    ## Setting the RNG seed to obtain consistent results.

## Generating a random DNA sequence
isq <- drawDNASequence(NN=length(er), piGap=0.2, pi=c(0.35,0.15,0.15,0.35))
rawToChar(isq)        ## Showing the random DNA sequence
```

where `piGap` is $\pi_{\text -}$ and `pi` contains $\pi_\mathrm{A}$,
$\pi_\mathrm{G}$, $\pi_\mathrm{C}$, and $\pi_\mathrm{T}$. The random DNA
segment drawn is therefore slightly AT-rich. Here, we will simulate the
evolution of this initial sequence assuming that all sites share the same shift
rate matrix $\mathbf{Q^*}$, but that each site has its own value of the mean
shift rate ($\rho$). To do so, we need to create a list of molecular evolution
simulators as follows:

```{r}
nesl <- list()              ## A list of nucleotide evolution simulators

## Initialize each simulator with a different mean shift rate:
for(j in 1:length(er))
  nesl[[j]] <- molEvolSim(Q, step = 1, rho = er[j])
```

To evolve the sequence, from its initial state using these molecular evolution
simulators, we call each simulator's function `$evolve()` function on its site
as follows:

```{r}
isq2 <- raw(length(isq))                  ## For storing the second sequence
for(j in 1:length(isq))                   ## For each site j
  isq2[j] <- nesl[[j]]$evolve(isq[j])     ## Evolve the site
rawToChar(isq)                            ## Show the first sequence
rawToChar(isq2)                           ## Show the second sequence
```

To evolve the sequence for even more steps, the time steps are themselves
iterated into a second loop as follows:

```{r, fig.width = 7.0}
stbl <- matrix(raw(),51,length(er))           ## For storing 51 sequences
stbl[1,] <- isq                               ## Initial sequence
stbl[2,] <- isq2                              ## Second sequence

## Simulation loops from squence 3 to 51:
for(i in 3:nrow(stbl))                              ## Loop for each sequence i
  for(j in 1:ncol(stbl))                            ## Loop for each site j
    stbl[i,j] <- nesl[[j]]$evolve(stbl[i - 1,j])    ## Simulate the shifts

## Showing the DNA sequences:
stbl %>%
  concatenate %>%
  show.sequence
```

In this code example, we introduced two more functions from **MPSEM**,
namely `concatenate()` and `show.sequence()`. Functions `concatenate()` calls
`rawToChar()` row by row on a `raw` matrix while optionally discarding a list
of characters (most generally the gap; ``"-"`), whereas `show.sequence()` is a
graphical function for showing sets of DNA sequences. For instance, one may
display the sequences while removing the gaps as follows:

```{r, fig.width = 7.0}
## Display the DNA sequences without the gaps:
stbl %>%
  concatenate(discard="-") %>%
  show.sequence
```

This is how sequences obtained from DNA sequencing would normally appear. Having
a set of sequences aligned by construct comes in handy when assessing the
performance of DNA sequence alignment algorithm and software. When necessary,
the sequences may be exported as a FASTA formal text file using another
**MPSEM** package function called `write.fasta()` as follows:

```{r}
stbl %>%
  concatenate(discard="-") %>%
  write.fasta(file="Test.fasta")
```

This file can be accessed by other software, for instance, `R` package **ape**:

```{r}
if(require(ape)) {
  tmp <- read.dna(file="Test.fasta", format="fasta", as.character=TRUE)
  tmp[1:5]
}
  
```

among others.

```{r, echo=FALSE, results='hide'}
file.remove("Test.fasta")
rm(tmp)
```

## Trait evolution simulator

**MPSEM** simulates quantitative traits as an Ornstein-Uhlenbeck process. The
Ornstein-Uhlenbeck process is analogous to a Brownian motion (diffusion)
process, but involves the mean and variance of the random fluctuations being
conditioned by the difference between the value and an optimum value in such a
way as to constrain the value to fluctuate more or less closely about that
optimum. The process is described formally as follows:

$$
x_t \sim \mathcal{N} \left(x_0 e^{-\alpha t} + (1 - e^{-\alpha t}) x_{opt}, \sigma \sqrt{\frac{(1 - e^{-2\alpha t})}{2 \alpha}} \right),
$$

where $x_0$ and $x_t$ are the trait value at time $0$ and after an amount of
time $t$, respectively, $x_{opt}$ is the optimal trait value, $\alpha$ is the
selection rate, and $\sigma$ is the standard deviation of a Gaussian process
denoted as $\mathcal{N}(\mu,\sigma)$ (where $\mu$ is the mean). When
$\alpha = 0$ the process becomes $x_t \sim \mathcal{N}(x_0,\sigma)$ as the
process becomes purely diffusive with a rate described solely by $\sigma$ and no
optimum exist.

In **MPSEM**, trait evolution is simulated using function `traitEvolSim()`.
Minimally, it needs three arguments as is called as follows:

```{r}
traitEvolSim(
    name = "Neutral",   ## Name of the trait
    sigma = 2.5,        ## Standard deviation (e.g., Mean leaf width in mm/Ma)
    step = 0.05         ## Simulation time step (e.g. Ma)
  ) -> ts_n
```

When called in this way, $\alpha = 0$ and the trait evolves according to a
purely diffusive process. The trait evolution is simulated as follows:

```{r, fig.width = 7.0}
trait <- numeric(1001)      ## For storing the values
trait[1] <- 40              ## Initial trait value

## Update the trait value at time t using the value at time t - 1:
for(i in 2:length(trait))
  trait[i] <- ts_n$updateValue(trait[i - 1])

plot(y = trait, x=0.05*(0:1000), type="l", las=1, ylab="Mean leaf width (mm)",
     xlab="Time (Ma)")
```

When argument $\alpha > 0$ (function call with `alpha = 1`), argument `optima`
needs to be provided value(s) as follows:

```{r}
traitEvolSim(
    name = "OUsingle",    ## Name of the trait
    alpha = 1,            ## The selection rate
    sigma = 2.5,          ## Standard deviation (e.g., Mean leaf width in mm/Ma)
    optima = 10,          ## A single trait optimum
    step = 0.05           ## Simulation time step (e.g. Ma)
  ) -> ts_OUs
```

Now, the simulator has a single optimum value of $10\,\mathrm{mm}$. Calculations
are performed in a similar fashion as follows:

```{r, fig.width = 7.0}
trait <- numeric(1001)      ## For storing the values
trait[1] <- 20              ## Initial trait value

## Update the trait value at time t using the value at time t - 1:
for(i in 2:length(trait))
  trait[i] <- ts_OUs$updateValue(trait[i - 1])

plot(y = trait, x=0.05*(0:1000), type="l", las=1, ylab="Mean leaf width (mm)",
     xlab="Time (Ma)")
abline(h=10, lty=3)
```

When more than one optima are provided through argument `optima`, a shift
intensity matrix must also be provided using argument `transition` as follows:

```{r}
traitEvolSim(
    name = "OUmultiple",  ## Name of the trait
    alpha = 1,            ## The selection rate
    sigma = 2.5,          ## Standard deviation (e.g., Mean leaf width in mm/Ma)
    optima = c(10,15,20), ## Three trait optima
    ## The trait optima shift intensity matrix:
    transition = matrix(c(NA,0.01,0,0.01,NA,0.01,0,0.01,NA), 3, 3),
    step = 0.05           ## Simulation time step (e.g. Ma)
  ) -> ts_OUm
```

Calculation involves keeping a value (or sets of values) of the state index,
which is updated using member function `$updateState()`, and is provided to
member function `$updateValue()` as follows:

```{r, fig.width = 7.0}
trait <- numeric(1001)      ## For storing the values
trait[1] <- 20              ## Initial trait value
state <- integer(1001)      ## For storing the optima
state[1] <- 1               ## Initial trait optimum

## Update the optimum at time t using the value at time t - 1, then
## update the trait value at time t using the value at time t - 1:
for(i in 2:length(trait)) {
  state[i] <- ts_OUm$updateState(state[i - 1])
  trait[i] <- ts_OUm$updateValue(trait[i - 1], state[i])
}

plot(y = trait, x=0.05*(0:1000), type="l", las=1, ylab="Mean leaf width (mm)",
     xlab="Time (Ma)")
lines(x=0.05*(0:1000), y=c(10,15,20)[state], col="black", lty=3L)
```

### Semi-quantitative or qualitative traits

Whereas these function simulate the evolution of quantitative traits, it is
straightforward to transform them into semi-quantitative or qualitative traits.
For instance, one may apply thresholds to transform the quantitative variables
into a semi-quantitative variables. To generate a binary variables (e.g., that
absence or presence of a feature), one can choose a threshold value at, and
above, which the trait is a presence and below which the trait is an absence.
For an $n$-class qualitative trait, $N$ deviates can be generated, one for each
of the classes, with deviate with the maximum value being giving the state of
the trait. The following code exemplifies the generation of an ordered, a
binary, and a five-category qualitative trait in `R` using simulated
quantitative traits:

```{r, fig.width = 7.0}
## Generate five random deviates, (n = 300 each) using a neutral evolution
## simulator:
traits <- matrix(NA, 300, 5)
traits[1,] <- c(0,0,1,0,0)

for(i in 1:5)
  for(j in 2:nrow(traits))
    traits[j,i] <- ts_n$updateValue(traits[j - 1,i])

## Simulate an ordered trait:
traits[,1] %>%
  sapply(
    function(x,b) {
      i <- sum(x > b)
      ifelse(i, i, 1)
    },
    b = quantile(.,seq(0,1,length.out=6))   ## For five classes
  ) %>%
  ordered -> var_ordered
levels(var_ordered) <- c("tiny","small","normal","large","huge")
pp <- par(no.readonly = TRUE)
par(mar=c(5,7,2,2))
plot(y=var_ordered, x=seq(0,50,length.out=length(var_ordered)), type="l",
     yaxt="n", xlab="time (Ma)", ylab="Value\n")
axis(2, at=1:5, labels=levels(var_ordered), las=1)
par(pp)

## Simulate a binary trait:
traits[,1] %>%
  {. > median(.)} -> var_binary
pp <- par(no.readonly = TRUE)
par(mar=c(5,7,2,2))
plot(y=var_binary, x=seq(0,50,length.out=length(var_binary)), type="l",
     yaxt="n", xlab="time (Ma)", ylab="Value\n")
axis(2, at=c(0,1), labels=c("Absent","Present"), las=1)
par(pp)

## Simulate a qualitative trait:
traits %>%
  apply(
    1L,
    which.max
  ) %>%
  c("blue","green","red","yellow","purple")[.] %>%
  as.factor -> var_qualitative
pp <- par(no.readonly = TRUE)
par(mar=c(5,7,2,2))
plot(y=var_qualitative, x=seq(0,50,length.out=length(var_qualitative)),
     type="l", yaxt="n", xlab="time (Ma)", ylab="Value\n")
axis(2, at=1:5, labels=levels(var_qualitative), las=1)
par(pp)
```

## Evolutionary network simulation

Whereas simulator generator functions `molEvolSim()` and `traitEvolSim()` allows
us to simulate the evolution of DNA sequences and quantitative traits, we now
have to apply them to more or less complex evolutionary graphs. Function
`randomGraph()` is a random graph simulator that allows one to simulate
different types of graphs using a process of single vertex addition. Its first
argument, `NV` is the number of vertices in the graph to be generated. In a
nutshell `randomGraph()` proceeds from a single origin vertex by adding one
vertex at a time and connecting it with the other vertices of the graph
following different approaches. At its creation, any vertex (including the
origin vertex) is given a maximum allowable number of outgoing edges (i.e.,
children). That number may be a constant or a random number, and is generated by
a function passed as an argument called `NC`. Every time an edge from a given
vertex is created, that number is decremented by $1$ until it reaches $0$, at
which point that vertex is no longer allowed to have any new edge going out from
it. The number of incoming edges (i.e., parents) of a newly created vertex is
obtained from a function passed as argument `NP` and must minimally be $1$. A
function is also passed to `randomGraph()` as its argument `timestep` and
returns the evolutionary distance of any edge generated during the process. The
last function that needs to be passed to `randomGraph()` as its argument
`maxDist` returns the maximum distance allowed between the parents of a new
vertex. This number constrains lateral transfers to vertices that are not too
dissimilar, thereby enforcing a barrier to reticulated evolution. Function
`randomGraph()` also accepts an arbitrary number of named arguments (i.e.,
`...`) and will pass them to the functions given as arguments `NC`, `NP`,
`timestep`, and `maxDiss`. For simplicity, it is recommended that these
functions have a `...` argument in order to filter out any unneeded arguments,
in addition to its own arguments, ideally with default values. For instance,
a dichotomic tree is generated as follows:

```{r, fig.height = 5, fig.width=7.15}
randomGraph(
  NV = 50,
  NC = function(...) 2,
  NP = function(...) 1,
  timestep = function(ts_min, ts_max, ...) runif(1, ts_min, ts_max),
  maxDist = function(...) NULL,
  ts_min = 2,
  ts_max = 5,
  verbose = FALSE
) -> gr_dich

gr_dich

plot(gr_dich, cex.min=3, cex.lab=0.5)
```

Here, the timestep is taken from a uniform random distribution with minimum
value $2$ and maximum value $5$, whereas the maximum number of children per
vertex is set to $2$ and the maximum number of parents per vertex is set to $1$
(i.e., the graph is a tree). Both functions `NV` and `NP` are given arbitrary
arguments `ts_min` and `ts_max`, but none will use them, whereas `timestep()`
uses both of them to generate random edge distances. `maxDist()` also disregards 
`ts_min` and `ts_max` as it consistently returns `NULL`. Actually, since only a
single incoming edge is ever used by the simulation process `maxDist()` will
never be executed. Another way to generate a random tree is function `rtree`
from package **ape**. For instance:

```{r, fig.height = 5, fig.width=7.15}
rtree(25) %>%
  as.graph -> gr_rtree

plot(gr_rtree, cex.min=1, show.vertex.labels = FALSE)
```

Contrary to the tree generated using function `randomGraph`, The tree generated
by **ape**'s function `rtree` is strictly dichotomic. Finally, a random
reticulated graph is generated as follows:

```{r, fig.height = 5, fig.width=7.15}
randomGraph(
  NV = 50,
  NC = function(lambda_child, ...) 1 + rpois(1, lambda_child),
  NP = function(lambda_parent, ...) 1 + rpois(1, lambda_parent),
  timestep = function(ts_min, ts_max, ...) runif(1, ts_min, ts_max),
  maxDist = function(max_anc, ...) runif(1, 0, max_anc),
  lambda_child = 3.0,
  lambda_parent = 3.0,
  ts_min = 5,
  ts_max = 12,
  max_anc = 13,
  verbose = FALSE
) -> gr_ret

gr_ret

plot(gr_ret, cex.min=3, cex.lab=0.5)
```

Here, the numbers of children and parents in excess of $1$ are drawn from
Poisson distributions with means `lambda_child` and `lambda_parent`,
respectively, and maximum distance between parents is drawn from an uniform
random distribution with minimum value $0$ and maximum value `max_anc`. For
plotting, the reticulated graph is coerced into a tree using method `as.phylo`
and is plotted using black segments, whereas the edges discarded during the
coercion process are drawn over the tree using red arrows.

We can now simulate DNA sequences and trait evolution on these random graphs
functions `simulateSequence` and `simulateTrait`. These function will invoke
DNA sequence and trait evolution models over a graph, taking the edge distances
as the evolutionary time. Function `simulateSequence` is called as follows:

```{r, fig.width=7.15}
gr_ret %>%
  simulateSequence(
    Q = DNArate(model="K80", piGap=0.25, insertionRate=0.15, deletionRate=0.15),
    sqn = drawDNASequence(NN=200, piGap=0.25, pi=c(0.3,0.2,0.2,0.3)),
    rate = drawEvolRate(NN=200, gamma.shape=4, gamma.scale=4e-4),
    a = 1
  ) -> seq_ret

seq_ret %>%
  concatenate(discard="-") %>%
  show.sequence
```

This function as the following arguments:

`x`
: A graph-class object on which the simulation is carried out

`Q`
: A shift rate matrix generated by function `DNArate`

`sqn`
: An initial DNA sequence such as the ones generated by function
`drawDNASequence`.

`rate`
: A vector of mean evolution rate for each of the nucleotides that can be
obtained from `drawEvolRate`

`weighting`
: a weighting function that assigns the contributions of multiple parent
vertices from the edge distances (more on below)

`...`
: further arguments to be internally passed to the `weighting` function.

The `weighting` function takes the (edge) distances from the parent vertices and
return contribution weights of these parents on the descendant vertex. These
weights thus had to sum to $1$ in every cases. The default function is the
following:

$$
w_a(d_i) = \frac{d_i^{-a}}{\sum_{i=1}^n d_i^{-a}},
$$

where $a$ is a parameter that dictates whether all vertices have the same
weights ($a=0$, the default), weights proportional with the inverse distance
($a=1$), proportional with the inverse square distance ($a=2$), and so on. In
the example above, $a=1$ and vertex weights were proportional with the inverse
edge distance. Custom contribution weighting function can be used in the place
of this default.

Function `simulateTrait` is called as follows:

```{r, fig.height = 5, fig.width=7.15, eval=FALSE}
gr_ret %>%
  simulateTrait(
    tem = ts_n,
    value = 0,
    a = 1
  ) -> trait_ret_n

trait_ret_n$value

plot(gr_ret, y=trait_ret_n$value, bg=head(rainbow(1200), 1000), cex.min=2,
     cex.lab=0.3)
```

to apply neutral trait evolution model contained in object `trait_ret_n`,
whereas applying the single optimum Ornstein-Uhlenbeck trait evolution model
contained in object `trait_ret_OUs`, one would proceed as follows:

```{r, fig.height = 5, fig.width=7.15, eval=FALSE}
gr_ret %>%
  simulateTrait(
    tem = ts_OUs,
    value = 0,
    a = 1
  ) -> trait_ret_OUs

trait_ret_OUs$value

plot(gr_ret, y=trait_ret_OUs$value, bg=head(rainbow(1200), 1000), cex.min=2,
     cex.lab=0.3)
```

In these two figures, a color scale is used to show the simulated trait values.
The plot method can also represent values at the vertices using marker size,
which is suitable for small graphs (see the documentation for the details).









